---
title: "Introduction à la visualisation sous R avec le package ggplot2"
subtitle: "Introduction"
author: "Jean-Michel Marin et Benjamin Charlier"
institute: "CNRS -- IMAG (Montpellier, France)"
date: Novembre 2022
output: binb::metropolis
classoption: "aspectratio=169,10pt"
header-includes:  \usepackage{xcolor}
---

## Introduction

\footnotesize

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(iris,aes(x=Sepal.Width, y=Sepal.Length, colour=Species))+
  geom_point(size=0.2)+
  scale_colour_manual(values=c("magenta", "orange", "blue"))+
  geom_smooth(method="lm", colour="red", size=0.15)
```

## Introduction

- ```ggplot2``` visualization package for R 
  ```R
  install.packages("ggplot2")
  install.packages("ggthemes")
  ```

- ```ggplot2``` is a system for declaratively creating graphics. Allows representation and exploration of datasets 

- Learning curve is steep but... investment is quickly extremely profitable
  - See the **documentation** at \url{https://ggplot2.tidyverse.org/}
  - See gallery of examples \url{https://r-graph-gallery.com/ggplot2-package.html}, \url{https://exts.ggplot2.tidyverse.org/gallery/}, \url{http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html} ...
  

- Other related project : ```plotly``` \url{https://plotly.com/r/}

## Main principle of ggplot2 package

\textcolor{red}{\large \bf ggplot2 syntax is about adding successives layers} 

- First layer is the **graph canvas**
  - Importing the considered data set, and variables names to be plotted

- Second layer is **aesthetic mapping**
  - Choosing the type of graph you want to plot: scatterplot, boxplot, barplot...

- Then come the **refinement layers**
  - they will allow you to choose the colors, the axis scales, the legend options...


## Instantiate a ggplot2 graph

To instantiate the ggplot2 graph, we use ```ggplot()```together with the argument ```aes()```

- ggplot2 graphs always start with this code
```R
ggplot(dataset, aes(x=, y = )) 
```

## Instantiate a ggplot2 graph

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(x=disp,y=mpg)) 
```

## Choosing a plot type : geom_XXX

To define the type of plot : scatter plot, boxplot, barplot... 

- Add a ```+``` operator at the end of the first line (ie ggplot2 class constructor), and add a new line with the desired function:
   - ```geom_point()```: scatter plot
   - ```geom_boxplot()```: boxplot
   - ```geom_bar()```: barplot...

## Choosing a plot type : geom_XXX

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(x=disp, y=mpg))+
  geom_point(size=0.5) 
```

## Graphical options

Finaly, fine tuning of aestethic options may be setup by adding extra layers 

- axes sizes and scales :  ```scale_x_continuous(trans='log10')```, ```xlim```, ```expand_limits```, ...

- colours : ```scale_colour_manual()```

- axes labels, with the functions ```xlab()```, ```ylab()```, ...

- legend or caption, e.g. ```theme(legend.position,=”bottom”)```

## Graphical options

\footnotesize 
To transform the x-axis into log scale add a layer with the fucntion ```coord_trans()```

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(x=disp, y=mpg))+
  geom_point(size=0.5)+
  coord_trans(x="log10") 
```

## Scatterplot with ggplot2

\footnotesize
Assume we need to graph a scatter plot with iris dataset: with x-axis being ```Sepal.Width``` variable and y-axis being ```Sepal.Length```:

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length))+
  geom_point(size=0.5) 
```

## Scatterplot with ggplot2

The iris dataset is composed with 3 diferent iris species (variable ```Species```) 

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, colour=Species))+
  geom_point(size=0.5) 
```

## Scatterplot with ggplot2

To change default colours with your custom choice, use the function ```scale_colour_manual()```. Careful with red/green (color blind) and yellow (readibility)...

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris,aes(x=Sepal.Width, y=Sepal.Length, colour=Species))+
  geom_point(size=0.5)+
  scale_colour_manual(values=c("magenta", "orange", "blue")) 
```

## Scatterplot with ggplot2

To display different point types, we use the shape argument in ``aes()``.

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, colour=Species, shape=Species))+
  geom_point(size=0.8)+
  scale_colour_manual(values=c("magenta", "orange", "blue")) 
```

## Scatterplot with ggplot2

To manually set point types, we use ```scale_shape_manual()```

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris,aes(x=Sepal.Width, y=Sepal.Length, colour=Species, shape=Species))+
  geom_point(size=0.8)+
  scale_shape_manual(values=c(21, 23, 25))
```

## Scatterplot with ggplot2

We are able to change the size of points. For instance, let the size depends on the value of some variable

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, size=Petal.Length))+
  geom_point(alpha=0.7) 
```

## Scatterplot with ggplot2

To add a regression curve or line, use the function ```geom_smooth()```

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, colour=Species))+
  geom_point(size=0.5)+
  geom_smooth(colour="red") 
```

## Scatterplot with ggplot2

Useful trick: adding a (least square) regression line, using the argument ```method="lm"```

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, colour=Species))+
  geom_point(size=0.5)+
  geom_smooth(method="lm", colour="red") 
```

## Scatterplot with ggplot2

It is also possible to add a (least square) regression by species, using the argument ```group``` in ```aes()``` (first line)

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(x=Sepal.Width, y=Sepal.Length, colour=Species, group=Species))+
  geom_point(size=0.5)+
  geom_smooth(method="lm") 
```

## Boxplot with ggplot2

Assume we now need a boxplot of the ```Sepal.Length``` variable by species... 

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris,aes(y=Sepal.Length, x=Species))+
  geom_boxplot() 
```

## Boxplot with ggplot2

To change the colour of the boxes, we use the ```fill``` option (instead of ```colour``` as before), in function ```aes()```:

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(y=Sepal.Length, x=Species, fill=Species))+
  geom_boxplot() 
```

## Boxplot with ggplot2

To add the observed data points, we use the ```geom_jitter()``` layer:

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris,aes(y=Sepal.Length, x=Species, fill=Species, colour=Species))+
  geom_boxplot(alpha=0.5)+
  geom_jitter(width=0.25, size=0.5) 
```

## Barplots with ggplot2

Counting barplot may be plotted easily:

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
ggplot(mtcars,aes(x=as.factor(gear)))+
  geom_bar() 
```

## Barplots with ggplot2

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(as.factor(gear), fill=as.factor(gear)))+
  geom_bar()
```

## Barplots with ggplot2
        
```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(as.factor(gear), fill=as.factor(carb)))+
  geom_bar() 
```

## Barplots with ggplot2

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(as.factor(gear)))+
  geom_bar(aes(fill=as.factor(carb)), position="fill") 
```

## Barplots with ggplot2

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
ggplot(mtcars,aes(as.factor(gear),fill=as.factor(carb)))+
  geom_bar(position="dodge") 
```

## Barplots with ggplot2

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(as.factor(gear)))+
  geom_bar(aes(fill=as.factor(carb)), position="dodge")+
  scale_fill_manual(values=c("red", "yellow", "blue", "magenta", "orange", "white")) 
```


## Pie chart with ggplot2

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=2, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(mtcars, aes(x=factor(1), fill=as.factor(gear)))+
  geom_bar(width=1)+
  coord_polar("y")+
  labs(x="", fill="Gear")
```


## Facetting (a.k.a. subplots)

\footnotesize
Dividing a graph into subplots, according to the modalities of one or more variables

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
library(MASS)
ggplot(Melanoma, aes(y=thickness,x=time))+
  geom_point(size=0.2)+
  geom_smooth()+
  facet_wrap(~status) 
```

## Facetting (a.k.a. subplots)

Dividing a plot according to the values of 2 variables

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(Melanoma, aes(y=thickness, x=time))+
  geom_point(size=0.2)+
  geom_smooth(size=0.15)+
  facet_grid(sex~status) 
```

## Mapping 

\footnotesize

The option ```colour=Species``` in the ```aes()``` function is **inherited** in every subsequent layers...

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(y=Sepal.Length, x=Species, colour=Species))+
  geom_violin()+
  geom_jitter(width=0.25, size=0.4) 
```

## Mapping 

To fix the colour in a particular layer, we have to explicitely call the ```colour``` option in this layer

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(y=Sepal.Length,x=Species,colour=Species))+
  geom_violin()+
  geom_jitter(width=0.25, colour="magenta", alpha=0.5, size=0.4) 
```

## Themes

There is a variety of different graphical theme

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris, aes(y=Sepal.Length, x=Species, colour=Species))+
  geom_violin()+
  geom_jitter(width=0.25, size=0.4)+
  theme_classic() 
```

## Themes

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=3.5, message=FALSE, warning=FALSE}
library(ggthemes) 
ggplot(iris, aes(y=Sepal.Length, x=Species, colour=Species))+
  geom_violin()+
  geom_jitter(width=0.25)+
  theme_excel()
```

## Axes, title and legend

```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
ggplot(iris,aes(y=Sepal.Length, x=Species, fill=Species,colour=Species))+
  geom_boxplot(alpha=0.5)+
  geom_jitter(width=0.25)+ggtitle("A veeeeeeeeery looooong tiiiiiiiiiitle")+
  ylab("Sepal Length")+
  xlab("Species")
```

## Outro

Consider ggplot2 extensions like ```animate``` : \url{https://gganimate.com/#yet-another-example}

<!-- ```{r echo=TRUE, cache=TRUE, fig.align='center', fig.height=2.5, fig.width=5.5, message=FALSE, warning=FALSE} -->

<!-- library(gapminder) -->
<!-- library(gganimate) -->

<!-- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) + -->
<!--   geom_point(alpha = 0.7, show.legend = FALSE) + -->
<!--   scale_colour_manual(values = country_colors) + -->
<!--   scale_size(range = c(2, 12)) + -->
<!--   scale_x_log10() + -->
<!--   facet_wrap(~continent) + -->
<!--   # Here comes the gganimate specific bits -->
<!--   labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') + -->
<!--   transition_time(year) + -->
<!--   ease_aes('linear') -->
<!-- ``` -->

## Exercise 1

With the ```mtcars``` dataset (from the R  package ```datasets```)

1. Plot an histogram of ```mpg``` variable
2. Plot a barplot of variable ```cyl```
3. Plot the point cloud ```disp``` vs ```mpg``` with a different colour depending on the  values of ```cyl``` variable 
```{r echo=FALSE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}
nb_cyl <- as.factor(mtcars$cyl)
ggplot(mtcars,aes(x=disp,y=mpg, colour=nb_cyl))+
geom_point()
```
## Exercise 2

1. Use ggplot2 to plot the sine function on the $[−2\pi, 2\pi]$.
2. Add the lines $y = −1$ and $y = 1$ in thick blue.
3. Add the cosine function.
4. Add legend to identify the sine and cosine function.
5. Plot cosine and sine on two subgraphs (use ```facet_wrap```).
```{r echo=FALSE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}

X <- seq(-2*pi,2*pi,by=0.1)
n <- length(X)
D <- data.frame(x=rep(X,2),
type=c(rep("sine",n),rep("cosine",n)),f=c(sin(X),cos(X)))


ggplot(D,aes(x=x,y=f,colour=type))+
  geom_line()+
  geom_hline(yintercept=-1,size=2, colour="blue")+
  geom_hline(yintercept=1,size=2, colour="blue")+
  theme(legend.position="none")+
  facet_wrap(~type)
```

## Exercise 3

1. Simulate a sample $(x_i , y_i ), i = 1, \cdots, 100$, following the model
$$Y_i = 3 + X_i + \varepsilon_i$$
where $X_i$ are i.i.d. uniform on $[0, 1]$ random variables and  $\varepsilon_i$ are i.i.d. centered Gaussian of variance $0.04$.

2. Plot the corresponding dataset as a point cloud and the least square regression line with the ```geom_smooth``` function.

3. Plot the residuals: add a vertical segment linking each point to the least square regression line.

```{r echo=FALSE, cache=TRUE, fig.align='center', fig.height=1.5, fig.width=5.5, message=FALSE, warning=FALSE}

n <- 100
x <- runif(n)
y <- 3+x+rnorm(n,sd=0.2)
D <- data.frame(x=x,y=y)

model.reg <- lm(y~x,data=D)

D <- data.frame(x=x, y=y, yhat=model.reg$fitted.values)

ggplot(D,aes(x=x,y=y))+
  geom_point()+
  geom_segment(data=D, aes(x=x, y=y, xend=x, yend=yhat), color="red", alpha=0.5)+
  geom_abline(intercept=model.reg$coefficients[1],slope=model.reg$coefficients[2], color="blue")+
  theme_classic()

```

## Exercise 4

Consider the dataset ```states``` from datasets package:
```R
data(state)
states <- data.frame(state.x77,state.name=rownames(state.x77), 
                     state.region=state.region)
```
1. Create a variable ```Income1``` taking the value ```low``` if the ```Income``` is in the first tercile, ```medium``` in the second and ```high``` in the third one (use functions ```quantile``` and ```cut```).
2. Plot the point clouds  ```Population``` vs ```Murder``` for each value of ```Income1``` (3 point clouds).
3. Use a different colour for each point according to the variable ```state.region``` and add the least square regression line on each graph.

## Exercise 4

```{r echo=FALSE, cache=TRUE, fig.align='center', fig.height=2.5, fig.width=6, message=FALSE, warning=FALSE}


data(state)
states <- data.frame(state.x77,state.name=rownames(state.x77), state.region=state.region)

# 1

q1 <- quantile(states$Income,1/3,names=FALSE)
q2 <- quantile(states$Income,2/3,names=FALSE) 
states$revenu1 <- cut(states$Income,breaks=c(min(states$Income),q1,q2,max(states$Income)),labels=c("low","medium","high"),include.lowest=TRUE)
# summary(states$revenu1)

# 2

# ggplot(states,aes(x=Population,y=Murder))+
# geom_point()+
# facet_wrap(~revenu1)+
# theme_classic()

# 3

ggplot(states,aes(x=Population,y=Murder))+
  geom_smooth(method="lm",se=FALSE)+
  geom_point(aes(colour=state.region), size=0.3)+
  facet_wrap(~revenu1)+
  theme(axis.text.x = element_text(angle = 30))
```

## Bibliography

- P.-A. Cornillon, A. Guyader, F. Husson, N. Jégou, J. Josse, N. Klutchnikoff, E. Le Pennec, E. Matzner-Løber, L. Rouvière, B. Thieurme - \textit{R pour la statistique et la science des données}. Pratique de la statistique, Presses universitaires de Rennes. 2018

- ggplot2 documentation. \url{https://ggplot2.tidyverse.org/}